Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-02-07 docs/ACM/paper.tex
Adding changes up to 3.4

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

@@

Figure~\ref{fig:ex_Regions} illustrates Challenge~\ref{ch:Regions}, and shows
how we solve it.
That program receives an $\mathsf{M}\times\mathsf{N}$ matrix \textsf{V}, in
linearized format, and produces a vector \textsf{U}, so that \textsf{U[i]}
contains the sum of all the elements in line \textsf{i} of matrix \textsf{V}.
For reasons to be considered in Section~\ref{sub:identification},
our static analysis determines that each iteration of the outermost loops
could be made into a task.
Thus, tasks comprise the innermost loop, and traverse the memory region
between addresses $\textsf{\&V + i * N}$ and $\textsf{\&V + i * N + M}$.
The identification of such ranges involves the use of a symbolic algebra,
which we have borrowed from the compiler-related literature, as we
explain in Section~\ref{sub:symb}.
Figure~\ref{fig:ex_Regions} also introduces the second challenge that we
tackle:

@@

Figure~\ref{fig:alg_main} provides a high-level view of a source-to-source
compiler that incorporates the techniques discussed in this paper.
This pseudo-code uses several concepts well-known in the compiler's
literature, such as control flow graphs~\cite{Kildall73} and
dependence graphs~\cite{Ferrante87}.
We built on top of such concepts, as well as others, like
symbolic range analysis~\cite{Blume94}, and natural loops~\cite{Allen70}.
In the rest of this section we explain how we have combined this previous
knowledge in order to delimit and annotate tasks in structured programs.
Our presentation focus on the new elements that we had to add onto known
techniques, in order to adapt them to our purposes.

@@

\paragraph{Tasks: syntax and scope:}
Our goal is to identify {\em tasks} in {\em structured} programs.
Definition~\ref{def:task} formalizes the notion of task.
A structured program is a program that can be partitioned in {\em hammock
regions}, a concept introduced by Ferrante {\em et al.}~\cite{Ferrante87} in
the middle 80's.
For completeness, we re-state it in Definition~\ref{def:hammock}.
The structure of this program is described by a a Control Flow Graph $G$ --
a directed graph with a starting node $s$, and an exit node $x$.

\begin{definition}[Hammock Region~\cite{Ferrante87}]
\label{def:hammock}
A hammock region $G'$ from a CFG $G$ is a subgraph $G' \subseteq G$, with a
node $h$, that {\em dominates}\footnote{A node $n_1 \in G$ dominates another node
$n_2 \in G$ if every path from  $S$ to $n_2$ goes across $n_1$.
Inversely, $n_1$ post-dominates $n_2$ if every path from $n_2$ to $x$ must
go across $n_1$.} the other nodes in $G'$.
Additionally, there exists a node $w \in G$, $w \notin G'$, such that
$w$ {\em post-dominates} every node in $G'$.
In this definition, $h$ is the {\em entry point}, and $w$ is the {\em exit point}
of $G'$.
\end{definition}

@@

\begin{example}[Memory Regions]
\label{ex:regions}
Below we have two tasks, $T_{\mathit{foo}} = (\mathsf{foo}, \{\mathsf{v[i - 1]}\},
\{v[i]\})$ and $T_{\mathit{bar}} = (\mathsf{bar}, \{v[i]\}, \{\})$.
Dependencies are identified via the \textsf{depend} clause.
In this example, each SESE region is formed by one function call:
\end{example}

@@

\paragraph{The annotation zoo.}
Our equipment of choice to reveal task parallelism in programs is OpenMP 4.0.
This annotation system provides us with a set of pragmas that indicate to compiler
and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

\begin{definition}[Windmill -- \cite{Rideau08}]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$, its center, plus $n$ components (not
necessarily strong) $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices.
\end{compactenum}
\end{definition}

@@

Windmills gives us structural properties to identify tasks.
We can easily find them via a depth-search traversal of the program dependence
graph, for instance.
However, not every windmill lead us to profitable tasks.
Furthermore, some windmills simply cannot be annotated, due to structural 
incompatibilities between the program dependence graph and its source code.
We deal with the first shortcoming in Section~\ref{sub:profit}; the latter is
the subject of Section~\ref{sub:ir}.

\subsection{Finding Symbolic Bounds of Arrays}
\label{sub:symb}

% Why this is necessary
To produce the annotations that will lead to the creation of a task $T = (G, M_i,
M_o)$, we need to determine the memory regions $M_i$ that $T$ reads, and the
memory regions $M_o$ that it writes.
Memory regions consist of pointers, plus ranges of addresses that these pointers
can dereference.
Example~\ref{ex:mem_reg} illustrates this notion.
To determine precise bounds to memory regions, we resort to an old ally of
compiler writers: the {\em Symbolic Range Analysis}, a concept that
Definition~\ref{def:sra} formalizes.

\begin{example}
\label{ex:sra}
Symbolic range analysis, when applied onto Figure~\ref{fig:sra}, gives us
$R(\mathsf{i}) = [0, \mathsf{N} - 1]$, and $R(\mathsf{j}) = [0, \mathsf{M} - 1]$.
The memory access \textsf{V[i*N + j]} at line 9, when combined with this
information, yields the symbolic region that appears in the task annotation at
line 6 of Figure~\ref{fig:ex_Regions}.
\end{example}

@@

% Explain the tension between large and small tasks.
The benefit of creating tasks depend on two factors:
runtime cost, and parallelism.
These two factors lead to a sweet spot in the size of tasks, where size is
measured as the amount of code they contain.
The smaller the tasks are, the more parallel they tend to be, as a consequence
of less dependencies.
However, if tasks are too small, then the performance gained by the extra
parallelism might not compensate the cost of managing them.
On the other hand, if tasks are too big, then we might not obtain enough
parallelism, because individual tasks still execute sequentially.
In this paper, {\em we extract the smallest thread above the runtime cost}.
To bestow meaning into this last statement, in what follows, we define more
formally the notions of task size and task expansion.

@@

Definition~\ref{def:swe} is too unconstrained: a function $W(G) = 0$ for any $G$
satisfies it.
Naturally, we would like to construct $W$ in such a way to approximate the 
dynamic behavior of programs.
The compiler-related literature contains heuristics to implement $W$.
Perhaps, the best-known among these heuristics is the static profiler of
Wu and Larus~\cite{Wu94}.
In this paper, we adopted a different approach: we reuse the symbolic range
analysis of Section~\ref{sub:sra} to augment static data with dynamic information.
In other words, we use symbolic range analysis to construct expressions that
represent the number of iterations of loops.
Similar approach has already been used to decide when to migrate virtual memory
pages in NUMA architectures~\cite{Piccoli14}.
OpenMP 4.0 contains syntax to enable the conditional creation of tasks.
Our symbolic analysis lets us build predicates for such conditional expressions.

\begin{example}
\label{ex:cond_task}
The condition in line 7 of Figure~\ref{fig:ex_Regions} determines that tasks will
only be created if the expression \textsf{5 $\times$ M $<$ WORK\_CUTOFF}
holds.
The expression \textsf{5 $\times$ M} is an approximation for the size of the
task that comprises the loop at lines 8-10.
The constant \textsf{WORK\_CUTOFF} represents the cost of creating a thread
in the OpenMP runtime, and it is determined empirically.
\end{example}

@@

% Explain how we use symbolic information to compute costs.
The declarative rules in Figure~\ref{fig:swe} sketches the heuristics that
we use to compute the SWE for a given hammock region $S$.
The only rules that deserves comments are \textsc{LoopInf} and
\textsc{LoopExp}.
The former applies on loops whose number of iterations we cannot bound via some
symbolic expression computed statically.
The latter represents loops that we can analyze using our Symbolic Range Analysis.
The auxiliary function $X(S)$ returns a symbolic expression that represents
the range of values covered by the code $S$ that controls the number of
iterations of the loop.

\pedro{We need to discuss the THRESHOLD. Are we going to use
César's experiments at Unicamp concerning the different runtimes? 
Because if yes we need to think on how to write it.}

% Discuss the expansion algorithm.
\paragraph{The Task Expansion Algorithm.}
Task expansion consists in finding the smallest task that is large enough to
pay for the thread creation cost.
Figure \ref{fig:expand_alg} shows the algorithm that performs this activity.
We start this algorithm by setting \textsf{REGION} to be a hammock graph $H$
that is contained entirely within a windmill $W$.
The hammock decomposition of a structured program forms a tree~\cite{Ferrante87},
such that vertices in this tree represent hammock graphs.
$H_1$ is a child of $H_2$ in this tree if $H_1 \subset H_2$, and for any other
node $H_x$, $H_1 \subset H_x \Rightarrow H_2 \subset H_x$.
In this context, we call $H_2$ the \textsf{PARENT} of $H_1$.

>>>>

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:sra}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

@@

Figure~\ref{fig:ex_Regions} illustrates Challenge~\ref{ch:Regions}, and shows
how we solve it.
That program receives an $\mathsf{M}\times\mathsf{N}$ matrix \textsf{V}, in
linearized format, and produces a vector \textsf{U}, so that \textsf{U[i]}
contains the sum of all the elements in line \textsf{i} of matrix \textsf{V}.
For reasons to be considered in Section~\ref{sub:identification},
our static analysis determines that each iteration of the outermost loops
could be made into a task.
Thus, tasks comprise the innermost loop, and traverse the memory region
between addresses $\textsf{\&V + i * N}$ and $\textsf{\&V + i * N + M}$.
The identification of such ranges involves the use of a symbolic algebra,
which we have borrowed from the compiler-related literature, as we
explain in Section~\ref{sub:sra}.
Figure~\ref{fig:ex_Regions} also introduces the second challenge that we
tackle:

@@

Figure~\ref{fig:alg_main} provides a high-level view of a source-to-source
compiler that incorporates the techniques discussed in this paper.
The many parts of this algorithm shall be explained throughout this section.
This pseudo-code uses several concepts well-known in the compiler's
literature, such as control flow graphs~\cite{Kildall73} and
dependence graphs~\cite{Ferrante87}.
In the rest of this section we explain how we have combined this previous
knowledge to delimit and annotate tasks in structured programs.
Our presentation focus on the new elements that we had to add onto known
techniques, in order to adapt them to our purposes.

@@

\paragraph{Tasks: syntax and scope:}
Our goal is to identify {\em tasks} in {\em structured} programs.
A structured program is a program that can be partitioned in {\em hammock
regions}, a concept introduced by Ferrante {\em et al.}~\cite{Ferrante87} in
the middle 80's.
For completeness, we re-state it in Definition~\ref{def:hammock}.
Definition~\ref{def:task} formalizes the notion of task.

\begin{definition}[Hammock Region~\cite{Ferrante87}]
\label{def:hammock}
A Control Flow Graph $G$ is a directed graph with a starting node $s$, and an
exit node $x$.
A hammock region $G$ is a subgraph of $G$ with a node $h$ that
{\em dominates}\footnote{A node $n_1 \in G$ dominates another node
$n_2 \in G$ if every path from  $S$ to $n_2$ goes across $n_1$.
Inversely, $n_1$ post-dominates $n_2$ if every path from $n_2$ to $x$ must
go across $n_1$.} the other nodes in $G'$.
Additionally, there exists a node $w \in G$, $w \notin G'$, such that
$w$ {\em post-dominates} every node in $G'$.
In this definition, $h$ is the {\em entry point}, and $w$ is the {\em exit point}
of $G'$.
\end{definition}

@@

\begin{example}[Memory Regions]
\label{ex:regions}
Below we have two tasks, $T_{\mathit{foo}} = (\mathsf{foo}, \{\mathsf{v[i - 1]}\},
\{v[i]\})$ and $T_{\mathit{bar}} = (\mathsf{bar}, \{v[i]\}, \{\})$.
Dependencies are identified via the \textsf{depend} clause.
In this example, each hammock region is formed by one function call:
\end{example}

@@

\paragraph{The annotation zoo.}
Our equipment of choice to reveal task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

\subsection{Finding Symbolic Bounds of Arrays}
\label{sub:sra}

% Why this is necessary
To produce the annotations that create a task $T = (G, M_i, M_o)$, we need to
determine the memory regions $M_i$ that $T$ reads, and the memory regions $M_o$
that it writes.
Memory regions consist of pointers, plus ranges of addresses that these pointers
can dereference.
Example~\ref{ex:mem_reg} illustrates this notion.
To determine precise bounds to memory regions, we resort to an old ally of
compiler writers: the {\em Symbolic Range Analysis}, a concept that
Definition~\ref{def:sra} formalizes.

\begin{example}
\label{ex:sra}
Symbolic range analysis, when applied onto Figure~\ref{fig:ex_Regions}, gives us
$R(\mathsf{i}) = [0, \mathsf{N} - 1]$, and $R(\mathsf{j}) = [0, \mathsf{M} - 1]$.
The memory access \textsf{V[i*N + j]} at line 9, when combined with this
information, yields the symbolic region that appears in the task annotation at
line 6 of Figure~\ref{fig:ex_Regions}.
\end{example}

@@

\begin{definition}[Windmill -- \cite{Rideau08}]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$, its center, plus $n$ components (not
necessarily strong) $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices (thus, sharing edges is also impossible).
\end{compactenum}
\end{definition}

@@

Windmills gives us structural properties to identify tasks.
We can find them via a depth-search traversal of the program's dependence
graph.
Function \textsf{findVanes}, in Figure~\ref{fig:alg_main} implements this
search.
However, not every windmill lead us to profitable tasks.
Furthermore, some windmills cannot be annotated, due to the inability of
range analysis to find bounds to the memory they access.
We deal with these shortcomings in the next two sections.

@@

% Explain the tension between large and small tasks.
The benefit of creating tasks depend on two factors:
runtime cost, and parallelism.
This tension leads to a sweet spot in the size of tasks, where size is
measured as the amount of code they execute.
The smaller the tasks are, the more parallel they tend to be, as a consequence
of less dependencies.
However, if tasks are too small, then the performance gained by the extra
parallelism might not compensate the cost of managing them.
On the other hand, if tasks are too big, then we might not obtain enough
parallelism, because individual tasks still execute sequentially.
In this paper, we mark as tasks {\em the first maximal analyzable vane above the
runtime cost}.
To imbue this last statement with meaning, in what follows, we define more
formally the notions of task size.
We leave the concept of maximal analyzable vane for Section~\ref{sub:expansion}.

@@

Definition~\ref{def:swe} is too unconstrained: a function $W(G) = 0$ for any $G$
satisfies it.
Naturally, we would like $W$ to approximate the 
dynamic behavior of programs.
The literature contains heuristics to implement $W$.
Perhaps, the best-known among these heuristics is the static profiler of
Wu and Larus~\cite{Wu94}.
In this paper, we adopted a different approach: we reuse the symbolic range
analysis of Section~\ref{sub:sra} to augment static data with dynamic information.
In other words, we use symbolic range analysis to construct expressions that
represent the number of iterations of loops.
Similar approach has already been used to decide when to migrate virtual memory
pages in NUMA architectures~\cite{Piccoli14}.
OpenMP 4.0 contains syntax to enable the conditional creation of tasks.
Our symbolic analysis lets us build predicates for such conditional expressions.

\begin{example}
\label{ex:cond_task}
The condition in line 7 of Fig.~\ref{fig:ex_Regions} determines that tasks can be
created if \textsf{5 $\times$ M $<$ WORK\_CUTOFF}
holds.
The expression \textsf{5 $\times$ M} is an approximation for the size of the
task that comprises the loop at lines 8-10.
The constant \textsf{WORK\_CUTOFF} represents the cost of creating a thread
in the OpenMP runtime, and it is determined empirically.
\end{example}

@@

% Explain how we use symbolic information to compute costs.
The declarative rules in Figure~\ref{fig:swe} sketches the heuristics that
we use to compute the SWE for a given hammock region $S$.
The only rules that deserves comments are \textsc{LoopInf} and
\textsc{LoopExp}.
The former applies on loops whose number of iterations we cannot bound via some
symbolic expression computed statically.
The latter represents loops that we can analyze using our Symbolic Range Analysis.
The auxiliary function $X(S)$ returns a symbolic expression that represents
the range of values covered by the code $S$ that controls the number of
iterations of the loop.
In Figure~\ref{fig:alg_main}, these rules are implemented by function
\textsf{profitabilityAnalysis}.

\paragraph{The Cost Model.}
A static workload estimate is only meaningful in the context of a {\em cost
model}.
A cost model is a collection of parameters that determine the impact of the
underlying computer architecture onto the creation and management of tasks.
The literature contains examples of techniques used to build cost models,
be it analytically~\cite{Baghsorkhi10}, or empirically~\cite{Poesia17}.
The automatic construction of a cost model is not the goal of this paper.
Instead, the reader shall notice that Figure~\ref{fig:alg_main} receives a
cost model as a parameter.
For the experiments in Section~\ref{sec:eval}, we have determined a small
collection of constants related to the creation of threads in our target
architecture, including the value of \textsf{WORK\_CUTOFF}, seen in
Example~\ref{ex:cond_task}.

% Discuss the expansion algorithm.
\subsection{Task Expansion}
\label{sub:expansion}

Task expansion consists in finding the smallest task that is large enough to
pay for the thread creation cost.
Figure \ref{fig:expand_alg} shows the algorithm that performs this activity.
We start this algorithm by setting \textsf{REGION} to be a hammock graph $H$
that is contained entirely within a windmill $W$.
The hammock decomposition of a structured program forms a tree~\cite{Ferrante87},
such that vertices in this tree represent hammock graphs.
$H_1$ is a child of $H_2$ in this tree if $H_1 \subset H_2$, and for any other
node $H_x$, $H_1 \subset H_x \Rightarrow H_2 \subset H_x$.
In this context, we call $H_2$ the \textsf{PARENT} of $H_1$.
